---
title: "ML Modelling"
output: html_document
---

```{r}
library(purrr)
library(ggplot2)
library(tidyr)
library(dplyr)
```

The toxicity of a chemical substance is tested by exposing it to beetles in different concentrations. The data are given in the ﬁle `ch4exerc2.csv`. The following table shows the results:

```{r}
ch4exerc2
```

`numberExposed` is the number of beetles exposed to the corresponding concentration, `numberDied` is the number of beetles that died at that concentration.

1.  Three different models are used to estimate the inﬂuence of toxicity on the probability that a beetle dies given a certain concentration $x$. Let $Y$ be the binary response with $Y=1$ if a beetle dies at concentration $x$ with $\pi(x) = P(Y = 1 \mid x)$ and $Y = 0$ otherwise.

    (a) The probability is linked to the concentration $x$ through the logistic function

        $$
        \pi(x) = \frac{\exp(\alpha_1 + \beta_1x)}{1 + \exp(\alpha_1 + \beta_1x)}
        $$

    (b) The probability is linked through the probit function

        $$
        \pi(x) = \Phi(\alpha_2 + \beta_2x)
        $$

        where $\Phi(\cdot)$ is the distribution function of the normal distribution.

    (c) The probability is linked through the complementary log-log function

        $$
        \pi(x) = 1- \exp[-\exp(\alpha_3 + \beta_3x)]
        $$

    For all three models, determine the likelihood and log-likelihood given the above data. Find the ML estimates of the parameters $(\alpha_j, \beta_j), j=1,2,3$ using a generic optimisation function in $R$, e.g. the function `optim`.

    We can model this with binomial variable. Given $Y$ and $X$ the number of beetles died and the concentration respectively, we can model $P(X,Y)$ as $\binom{Y_{\mathtt{Exposed}}}{Y_{\mathtt{Died}}} \pi(x)^{Y_{Died}} (1 - \pi(x))^{Y_{\mathtt{Exposed}} - Y_{Died}}$.

    ```{r}
    pi_1 <- function(alpha, beta, x){
      exp(alpha + beta*x)/(1 + exp(alpha + beta*x))
    }

    pi_2 <- function(alpha, beta, x){
      dnorm(x = alpha + beta*x)
    }

    pi_3 <- function(alpha, beta, x){
      1 - exp(-exp(alpha + beta*x))
    }
    ```

    ```{r}
    binom_prob <- function(x, y_died, y_exp, prob_func, alpha, beta){
      dbinom(y_died, size = y_exp, prob = prob_func(alpha, beta, x))
    }

    likelihood <- function(alphabeta, dataset, prob_func){
      binom_before <- function(X, YD, YE) binom_prob(X, YD, YE, prob_func, alphabeta[1], alphabeta[2])
      binom_parallel <- pmap(list(X=dataset$concentration, 
                                  YD=dataset$numberDied, 
                                  YE=dataset$numberExposed), 
                                  binom_before) |> unlist(use.names = FALSE)
      prod(binom_parallel)
    }

    likelihood(c(0.5,0.01), ch4exerc2, pi_3)
    ```

    ```{r}
    likelihood_pi_1 <- function(alphabeta) (-1)*likelihood(alphabeta, ch4exerc2, pi_1) |> log()
    likelihood_pi_2 <- function(alphabeta) (-1)*likelihood(alphabeta, ch4exerc2, pi_2) |> log()
    likelihood_pi_3 <- function(alphabeta) (-1)*likelihood(alphabeta, ch4exerc2, pi_3) |> log()

    optim(par = c(0,0), fn = likelihood_pi_1)$par
    optim(par = c(0,0), fn = likelihood_pi_2)$par
    optim(par = c(0,0), fn = likelihood_pi_3)$par
    ```

2.  Using the Maximum Likelihood estimates $(\hat{\alpha}_j, \hat{\beta}_j),j=1,2,3$ calculate the expected proportion of dead beetles for each concentration. Compare the results with the raw proportions (`numberDied`/`numberExposed`) of the data and visualise the results in an appropriate plot.

    ```{r}
    x_pi <- seq(from = 1.5, to = 2.1, by = 0.01)
    y_pi1 <- pi_1(-59.83858, 33.77936, x_pi)
    y_pi2 <- pi_2(13.488121,-7.264653, x_pi)
    y_pi3 <- pi_3(-36.95845, 20.57205, x_pi)
    pi1_graph <- data.frame(x_pi, y_pi1, y_pi2, y_pi3)
    pi1_graph <- pi1_graph |>
      pivot_longer(cols = 2:4,
                   names_to = 'y_pi_j',
                   values_to = 'value')

    concent_sample <- ch4exerc2 |>
      mutate(pi_x_sample = numberDied/numberExposed)

    concent_estimate <- data.frame(concentration = ch4exerc2$concentration) |>
      mutate(y_pi1 = pi_1(-59.83858, 33.77936, concentration),
             y_pi2 = pi_2(13.488121,-7.264653, concentration),
             y_pi3 = pi_3(-36.95845, 20.57205, concentration)) |>
      pivot_longer(cols = c('y_pi1', 'y_pi2', 'y_pi3'),
                   names_to = 'y_pi_j',
                   values_to = 'value')
    ```

    ```{r}
    ggplot(pi1_graph, aes(x = x_pi, y = value)) +
      geom_line(aes(color = y_pi_j)) +
      geom_point(data = concent_sample, 
                 aes(x = concentration, y = pi_x_sample), 
                 size = 3, shape = 21) +
      geom_point(data = concent_estimate, 
                 aes(x = concentration, y = value, color = y_pi_j),
                 alpha = 0.5, size = 2) +
      theme_minimal()
    ```

    ```{r}
    ggplot(concent_estimate, aes(x = concentration, y = value)) +
      geom_vline(xintercept = concent_estimate$concentration,
                 alpha = 0.2) +
      geom_point(aes(color = y_pi_j, shape = y_pi_j),
                 size = 3) + 
      geom_point(data = concent_sample, aes(x = concentration, y = pi_x_sample),
                 size = 3, shape = 3) +
      theme_minimal()
    ```

3.  Think about how to determine which of the three models has the best ﬁt to the observed proportions.

    In terms of the model, logistic models is the most suitable model conceptually, especially model 1 and 3. In some way, a chemical substance will be lethal after a certain concentration. Plus, these models fit the sample data way quite so.
