---
title: "Central Limit Theorem for Poisson Distribution"
output: html_document
---

Write a simulation program using random numbers and plots to show that the central limit theorem holds for independent Poisson distributed random variables. Evaluate, whether the speed of convergence depends on the choice of the parameter $\lambda$: $\lambda = 0.1$, $\lambda = 0.5$, $\lambda = 1.0$, $\lambda = 5$, $\lambda = 10$, $\lambda = 50$. Discuss your results.

We will construct a data frame consisting of different columns on these parameters and convert them to their means, cumulatively calculated.

```{r}
num_entry = 1000
pois_lambda <- matrix(data = rep(x = 0, times = 6*num_entry), nrow = num_entry, ncol = 6)
mean_pois_lambda <- pois_lambda

lambda_const = c(0.1, 0.5, 1.0, 5.0, 10.0, 50.0)
lambda_char = as.character(lambda_const)
for(i in 1:num_entry){
  for(j in 1:6){
    pois_lambda[i,j] <- rpois(n=1, lambda = lambda_const[j])
    mean_pois_lambda[i,j] <- mean(pois_lambda[1:i,j])
  }
}

mean_pois_lambda <- as.data.frame(mean_pois_lambda)
colnames(mean_pois_lambda) <- lambda_char
mean_pois_lambda

```

Next is to plot the convergence.

```{r}
library(ggplot2)
library(dplyr)
library(tidyr)
library(ggforce)
```

```{r}
mean_pois <- within(data = mean_pois_lambda, expr = {
  n_sample <- 1:num_entry
}) |> gather(key = "lambda",
             value = "mean",
             convert = TRUE,
             c(-n_sample)) |>
  mutate(lambda_factor = factor(lambda))

glimpse(mean_pois)

for_hlines <- data.frame(lambda_factor = lambda_char, mean = lambda_const)

ggplot(data = mean_pois, aes(x = n_sample)) +
  geom_line(aes(y = mean, color = lambda_factor)) +
  facet_grid(rows = vars(lambda_factor), scales = 'free_y') +
  geom_hline(aes(color = lambda_factor, yintercept = mean),
             for_hlines, linetype = 2, alpha = 0.8) +
  theme_bw()

ggplot(data = mean_pois, aes(x = n_sample)) +
  geom_line(aes(y = (mean - lambda) |> abs(), 
                color = lambda_factor), 
            alpha = 0.7, linewidth = 0.8) +
  facet_zoom(xlim = c(750, 1000),
             ylim = c(0, 0.5)) +
  theme_bw()
```

As we can see, higher values of $\lambda$ lead to slower convergence. This is because higher moments of Poisson distribution depends on the variable on a proportional level, and our $K_{Z_n}(t)$ will need a bigger $n$ to diminish those terms in its Taylor series representation. That's why the convergence will be relatively slower.

However, with respect to its relative variance scale, that is, $\tfrac{Z_n-\mu}{\sigma} = \tfrac{Z_n-\lambda}{\sqrt \lambda}$, things might be different:

```{r}
ggplot(data = mean_pois, aes(x = n_sample)) +
  geom_line(aes(y = ((mean - lambda)/sqrt(lambda)) |> abs(), 
                color = lambda_factor), 
            alpha = 0.7, linewidth = 0.8) +
  facet_zoom(xlim = c(750, 1000),
             ylim = c(0, 0.1)) +
  theme_bw()
```

in that it seems like there's not any significance. Note that the distribution $\tfrac{Z_n-\lambda}{\sqrt \lambda}$ is essentially the distribution $N(0,1)$ when $n$ is large enough. When seen in this way, we can measure how relatively convergent it is. However there's not a pattern to be seen or we might have made the wrong judgement.
