---
title: "Trivariate Random Normal Variables"
output: html_document
---

Random numbers are often used to design simulation studies for evaluating the performance of statistical procedures. Generate a sample of $n=200$ trivariate normal random numbers with mean $\mu = (3.0,10.0,50.0)$ and covariance matrix

$$
\Sigma = \pmatrix{1.0&0.6&0.6\\0.6&1.0&0.6\\0.6&0.6&1.0}
$$

1.  As a ﬁrst method, use the fact that, if $X$ is a random vector of independent standard normal $N(0,1)$ variables, then

    $$
    Y= \mu + \Sigma^{1/2}X
    $$

    has the desired mean and covariance matrix. Use e.g. the Cholesky decomposition in R to compute the matrix root 1/2. By repeating the experiment $S=1000$ times, check empirically that your random numbers have the desired expected values, variances and covariances.

The Cholesky decomposition decomposes the positive real symmetric matrix $\Sigma$ into $\Sigma^{1/2}(\Sigma^{1/2})^T$ where $\Sigma^{1/2}$ is a lower/upper triangular matrix. The function `chol()` in R makes them upper triangular, so transposing is in order.

```{r}
mu <- c(3.0, 10.0, 50.0)
covar_matrix <- matrix(data = c(1.0, 0.6, 0.6, 0.6, 1.0, 0.6, 0.6, 0.6, 1),
                       nrow = 3, ncol = 3)
covar_matrix

# Calculate Cholesky decomposition
sqrt_covar_matrix <- chol(covar_matrix) |> t()
sqrt_covar_matrix

# Verify
sqrt_covar_matrix %*% t(sqrt_covar_matrix)
```

Do the experiment using the `rnorm` function.

```{r}
# Generate the random vectors
rand_vectors <- matrix(nrow = 3, ncol = 1000)
for(i in 1:1000){
  rand_vectors[,i] <- rnorm(n = 3, mean = 0, sd = 1)
}
rand_vectors[,1:5]

# Create an extension of mu
mu_matrix <- matrix(rep(x=mu,times=1000), nrow = 3, ncol = 1000)
mu_matrix[,1:5]

# Calculate Y
Y_rand_vectors <- mu_matrix + sqrt_covar_matrix %*% rand_vectors
Y_rand_vectors[,1:5]

# Calculate mean and variances:
print("Means:")
apply(Y_rand_vectors, MARGIN=1, FUN = mean)
print("Variances:")
apply(Y_rand_vectors, MARGIN=1, FUN = sd)
```

2.  A more comfortable way is to use the function `rmvnorm` in the package `mvtnorm`. Repeat your experiment using this function.

```{r}
library("mvtnorm")
```

```{r}
# rmvnorm
# n = number of observations
# mean = vector of means
# sigma = vector of covariances
# method = methods to be chosen (eigen, svd or chol)
rmvnorm(n=1000, mean = mu, sigma = covar_matrix, method = "chol")
```

3.  Calculate the (conditional) variance of the conditional distribution of $\left(Y_1 \mid Y_2,Y_3\right)$.

As outlined by the text, we divide our random vector $Y=(Y_1,Y_2,Y_3)^T$ into $(Y_A^T, Y_B^T)$ such that $Y_A = (Y_1)^T$ and $Y_B = (Y_2, Y_3)^T$. Our covariance matrix is now blocked into:

$$
\Sigma =
\left(\begin{array}{c:cc}
1.0&0.6&0.6\\ \hdashline 0.6 & 1 & 0.6\\ 0.6 & 0.6 & 1
\end{array}\right)
= \pmatrix{\Sigma_{AA}&\Sigma_{AB}\\\Sigma_{BA}&\Sigma_{BB}}\text{.}
$$

Thus we can calculate the mean with $\tilde{\mu} = \mu_A - \Sigma_{AB}\Sigma_{BB}^{-1}(y_B - \mu_B)$:

```{r}
install.packages("matlib")
library(matlib)
```

```{r}
covar_matrix[2:3,2:3]
```

```{r}
mu_A <- c(3.0)
sigma_AA <- covar_matrix[1,1]
sigma_AB <- covar_matrix[1,2:3]
sigma_BA <- covar_matrix[2:3,1]
sigma_BB <- covar_matrix[2:3,2:3]

coeff_ya_yb <- sigma_AB %*% (sigma_BB |> inv())
coeff_ya_yb
```

which gets us the expression

$$
\tilde{\mu} = 3.0 - \pmatrix{0.375&0.375}\pmatrix{y_2 -10\\y_3-50}
$$

and lastly, $\widetilde{\Sigma} = \Sigma_{AA} - \Sigma_{AB}\Sigma_{BB}^{-1}\Sigma_{BA}$ is calculated as follows:

```{r}
sigma_tilde <- sigma_AA - sigma_AB %*% (inv(sigma_BB)) %*% sigma_BA
sigma_tilde
```

thus $\widetilde{\Sigma} = 0.55$
