---
title: "Poisson Estimations"
output: html_document
---

In the data `injured.csv`, the weekly numbers of pedestrians severely injured in traffic in a city are recorded for 10 years (n = 520). We assume that the numbers are independent realizations of a Poisson distributed random variable $Y$ with constant intensity parameter $\lambda$.

1.  Derive and calculate the Maximum Likelihood estimate $\hat \lambda_{\text{ML}}$ of $\lambda$ given the available data.

Suppose we have samples $Y_1, \ldots, Y_n$ where $Y_i \sim \text{Pois}(\lambda)$ for all $i$. Given the data $y$, our Maximum Likelihood function would be

$$
L(y_1,\ldots,y_n ; \lambda) = \prod_{i=1}^{n} f(y_i;\lambda) = \prod_{i=1}^{n} \frac{e^{-\lambda}\lambda^{y_i}}{y_i!} = \frac{e^{-n\lambda}\lambda^{\sum y_i}}{\prod y_i!}
$$

Towards its log-likelihood, we have

$$
l(y_1,\ldots,y_n ; \lambda) =-n\lambda+(\Sigma_i y_i) \log \lambda - \log {\Pi_i \,y_i!}
$$

Differentiating and setting it to zero we have

$$
\begin{align*}
0 &= -n + {\Sigma_i y_i \over \hat \lambda_{\text{ML}}} \\
\hat \lambda_{\text{ML}} &= \frac{\Sigma_i y_i}{n}
\end{align*}
$$

which is supposedly the average of the data. By the following code, we have $\hat \lambda_{\text{ML}} = 4.996154$.

```{r}
mean(injured[["y"]], na.rm = TRUE)
```

2.  Calculate the method of moments estimate for $\lambda$

This one is quite easy to see because $\mathbb{E}[Y] = \lambda$, creating the equation $\hat \lambda_{\text{MM}} = m_1 = \Sigma_i y_i /n = 4.996154$.

3.  Calculate a $95\%$ confidence interval, assuming asymptotic normality of the ML estimate.

Asymptotic normality makes us deduce that

$$
{\hat \lambda_{\text{ML}} - \lambda \over \sqrt {\text{Var}(\hat\lambda_{\text{ML}})}} \overset a \sim N(0,1)
$$

when our number of samples goes high, which in turn pivots our calculation of the confidence interval on a normal distribution:

$$
\begin{align*}
1 - \alpha \approx \mathbb{P}(\hat\lambda_{\text{ML}} - z_{1-\alpha/2} \sqrt {\text{Var}(\hat\lambda_{\text{ML}})} \leq \lambda \leq \hat\lambda_{\text{ML}} + z_{1-\alpha/2} \sqrt {\text{Var}(\hat\lambda_{\text{ML}})})
\end{align*}
$$

where $z_{1 - \alpha/2}$ is the $1-\alpha/2$-th quantile of $N(0,1)$. But first we have to calculate the variance of $\hat\lambda_{\text{ML}}$ as what follows:

$$
\begin{align*}
\text{Var}(\hat\lambda_{\text{ML}}) &= \frac{1}{n^2}\Sigma_i \text{Var}(Y_i) = \frac{1}{n^2}(n\lambda) = \frac{\lambda}{n}\\
\implies\text{Var}(\hat\lambda_{\text{ML}}) &= \frac{\hat\lambda_{\text{ML}}}{n}
\end{align*}
$$

Since it only depends on $\lambda$, we can just replace it with $\hat \lambda_{\text{ML}}$. What we have to do now is calculate through R:

```{r}
alpha = 0.05
n = injured[["y"]] |> length()
z_quant <- qnorm(1 - alpha/2, mean = 0, sd = 1)
lambda_ml <- mean(injured[["y"]], na.rm = TRUE)
var_ml <- lambda_ml/n
lower_bound <- lambda_ml - z_quant*sqrt(var_ml)
upper_bound <- lambda_ml + z_quant*sqrt(var_ml)

print(paste0("[",lower_bound,", ",upper_bound,"]"))
```

Therefore, our confidence interval is

$$
\begin{align*}
\text{CI}_{95\%} &= \left[\hat\lambda_{\text{ML}} - z_{1-\alpha/2} \sqrt {\text{Var}(\hat\lambda_{\text{ML}})}, \hat\lambda_{\text{ML}} + z_{1-\alpha/2} \sqrt {\text{Var}(\hat\lambda_{\text{ML}})}\right] \\
&= \left[4.80403749888434, 5.18827019342335\right]
\end{align*}
$$
